Breaking the curse of dimensionality for machine learning on genomic data

### 2017-08-09 doc/publications/IJCAI2017/IJCAI.tex
tex update

\documentclass[a4paper]{amsart}
\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{epstopdf}
\usepackage{subcaption}

@@

\title{CursedForest - A Random Forest Implementation for ``Big'' and ``Wide'' Data}
%\section{}
%\subsection{}
\begin{document}  
\maketitle

\begin{abstract}
Genomic data analyses are performed on ever larger patient cohorts with unprecedentedly detailed genome-wide information for each individual. Machine learning (ML) is employed to detect the complex genomic interactions that can lead to diseases like diabetes or cancer. 
However, current ML approaches are unable to cope with these data volumes and even implementations using novel compute paradigms, such as Hadoop/Spark, are limited by the finite resource of computer memory. Specifically, Spark ML?s supervised machine learning methods suffer from "the curse of dimensionality". That is, they scale well with samples (n) but not with features (p, variants). 
We introduce CursedForest, a tailored implementation of random forests, designed to handle data with extremely large number of variables per sample. We showcase CursedForest to accurately predict ethnicity from whole genome sequencing profiles of 2,500 individuals and 80 million genomic variants each (1000 Genomes Project) in 30 minutes. CursedForest was the only implementation scaling to the full dataset, outperforming other methods on smaller subsets by 80\% or more. We also demonstrate a GWAS-style analysis using CursedForest?s feature selection approach on an imputed SNP array dataset (2000 x 7 Million) on Bone mineral density (BMD), successfully predicting known BMD-genes as well as identifying novel candidates. 
CursedForest is included in our earlier genome interpretation package, VariantSpark, allowing it to perform near real-time classification of population-scale patient cohorts in "patients like mine" scenarios, as well as performing GWAS analysis on large unfiltered whole genome sequencing cohorts. 
\end{abstract}

>>>>

\documentclass[a4paper]{amsart}
\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{epstopdf}
\usepackage{subcaption}q

@@

\title{Breaking the curse of dimensionality for machine learning on genomic data}
%\section{}
%\subsection{}
\begin{document}  
\maketitle

\begin{abstract}
Genomic data analyses are performed on ever larger patient cohorts with unprecedentedly detailed genome-wide information for each individual. Machine learning (ML) is employed to detect the complex genomic interactions that can lead to diseases like diabetes or cancer. 
However, current ML approaches are unable to cope with these data volumes and even implementations using novel compute paradigms, such as Hadoop/Spark, are limited by the finite resource of computer memory. Specifically, Spark ML's supervised machine learning methods suffer from "the curse of dimensionality". That is, they scale well with samples (n) but not with features (p, variants). 
We introduce CursedForest, a tailored implementation of random forests, designed to handle data with extremely large number of variables per sample. We showcase CursedForest to accurately predict ethnicity from whole genome sequencing profiles of 2,500 individuals and 80 million genomic variants each (1000 Genomes Project) in 30 minutes. CursedForest was the only implementation scaling to the full dataset, outperforming other methods on smaller subsets by 80\% or more. We also demonstrate a GWAS-style analysis using CursedForest's feature selection approach on an imputed SNP array dataset (2000 x 7 Million) on Bone mineral density (BMD), successfully predicting known BMD-genes as well as identifying novel candidates. 
CursedForest is included in our earlier genome interpretation package, VariantSpark, allowing it to perform near real-time classification of population-scale patient cohorts in "patients like mine" scenarios, as well as performing GWAS analysis on large unfiltered whole genome sequencing cohorts. 
\end{abstract}
